---
title: "Students' Grades Prediction Data Exploration"
output: html_notebook
---

The data used in this analysis is from https://www.kaggle.com/aljarah/xAPI-Edu-Data.

# attaching relevant libraries
```{r}
library(dplyr)
library(ggplot2)
library(gridExtra)
```

# Data Exploration

```{r}
grades <- read.csv("xAPI-Edu-Data.csv", stringsAsFactors = T)
grades$Class <- factor(grades$Class, levels = c("L", "M", "H"))
head(grades, 10)
```

```{r}
summary(grades)
```

```{r}
glimpse(grades)
```


```{r}
anyNA(grades)
```

There are no missing values in the dataframe.


Decided to choose Number of times students raised their hands (raisedhands), visited resources (VisITedResources), view announcements (AnnouncementsView), and take part in discussions (Discussion) to analyse against their grade category (Class). Reshaped the dataframe to more easily create data visualisations.
```{r}
library(reshape2)

grades2 <- grades %>%
  dplyr::select(raisedhands, VisITedResources, AnnouncementsView, Discussion, Class) %>%
  dplyr::rename(raised_hands = raisedhands, visited_resources = VisITedResources, announcements_viewed = AnnouncementsView, discussion = Discussion)
grades2_long <- reshape2::melt(grades2)
```

According to the boxplots, students with low grades (L) raised their hands in class, visitied resources, and view announcements 2 to 3 times lesser than those with medium (M) or high grades (H). Those who get high grades tend to raised their hands and visit resources more overall than those with medium grades.
```{r}
ggplot(grades2_long,aes(x= Class,y=value))+
  geom_boxplot(aes(fill=variable)) +
  ggtitle("Frequency of raised hands, visited resources, viewed announcements, & participate in discussions for each Grade") +
  theme(plot.title = element_text(size = 8, face="bold"))+
  xlab("Grades") +
  ylab("Frequency")
```
Their distribution are mainly bivariate but we won't know if it will negatively impact the model until we test it.

```{r}
#distribution of raisedhands
dist1 <- ggplot(data=grades2, aes(raisedhands)) + 
  geom_histogram(bins=30) +
  ggtitle("Distribution of raised hands")
```

```{r}
dist2 <- ggplot(data=grades2, aes(VisITedResources)) + 
  geom_histogram(bins=30) +
  ggtitle("Distribution of visited resources")
```

```{r}
dist3 <- ggplot(data=grades2, aes(AnnouncementsView)) + 
  geom_histogram(bins=30) +
  ggtitle("Distribution of view announcements")
```

```{r}
dist4 <- ggplot(data=grades2, aes(Discussion)) + 
  geom_histogram(bins=30) +
  ggtitle("Distribution of discussion participation")
```


```{r}
grid.arrange(dist1, dist2, dist3, dist4, nrow=2)
```

The classes are currently unbalanced.
```{r}
ggplot(grades, aes(fill=Class, x=Class)) +
  geom_bar() +
  ggtitle("Proportion of each Grade Class")
```

